---
title: "US Soccer Attendance"
author: "Connor Campbell"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse)
library(rvest)
library(fuzzyjoin)
library(scales)
library(ggthemes)
library(readxl)
library(here)
library(rio)
library(kableExtra)
```

# Introduction

This is an analysis of US professional soccer attendance data. Through
looking at both MLS and USL attendance trends for 2022 and 2023 I am
hoping to come to some insights about what is driving attendance in US
soccer and how the USL and MLS (the two main professional leagues)
compare.

# Data Collection

## Inputing and Cleaning USL and MLS Attendance Data

Utilizing rvest package I scraped data from soccerstadiumdigest.com to
get data for the USL Championship league. I then coerced the
imputed data into a data frame for easier manipulation and renamed a few
of the columns.

```{r}
html <- read_html("https://soccerstadiumdigest.com/2023-usl-championship-attendance/")

usl_2023_attendance <- html %>% 
    html_node("table") %>% 
    html_table(
      header = TRUE,
    )

usl_2023_attendance <- as.data.frame(usl_2023_attendance)
usl_2023_attendance <- usl_2023_attendance[,-1]
usl_2023_attendance <- usl_2023_attendance[,-7]
usl_2023_attendance <- usl_2023_attendance[-(24:26),]


usl_2023_attendance <- usl_2023_attendance %>% 
  rename(Average = Av.,
         Games = Gms,
         '2022 Average' = '2022')

```

Following this I then did some quick cleaning to get rid of commas and
convert some of the DF columns into numeric values.

```{r}
usl_2023_attendance$Average <- gsub(",","",usl_2023_attendance$Average)
usl_2023_attendance$Totals <- gsub(",","",usl_2023_attendance$Totals)
usl_2023_attendance$`2022 Average` <- gsub(",","",usl_2023_attendance$`2022 Average`)

usl <- usl_2023_attendance %>% 
  mutate_at(c('Average','Totals','2022 Average'), as.numeric)
```

Here is a view of the USL Championships 2023 attendance data.

```{r echo = FALSE}
usl %>%
  kbl() %>%
  kable_styling(fixed_thead = T)
```

Now following the same steps with the 2023 MLS data from
<https://soccerstadiumdigest.com/2023-mls-attendance/> we get this
result.

```{r include=FALSE}
htmlMLS <- read_html("https://soccerstadiumdigest.com/2023-mls-attendance/")

MLS_2023_attendance <- htmlMLS %>% 
  html_node("table") %>% 
  html_table(
    header = TRUE
  )

MLS_2023_attendance <- as.data.frame(MLS_2023_attendance)
MLS_2023_attendance <- MLS_2023_attendance[,-1]
MLS_2023_attendance <- MLS_2023_attendance[,-(7)]
MLS_2023_attendance <- MLS_2023_attendance[-(30:31),]

MLS_2023_attendance <- MLS_2023_attendance %>% 
  rename(Average = Av.,
         Games = Gms,
         '2022 Average' = '2022',
         Totals = Total)

MLS_2023_attendance$Average <- gsub(",","",MLS_2023_attendance$Average)
MLS_2023_attendance$Totals <- gsub(",","",MLS_2023_attendance$Totals)
MLS_2023_attendance$`2022 Average` <- gsub(",","",MLS_2023_attendance$`2022 Average`)

MLS <- MLS_2023_attendance %>% 
  mutate_at(c('Average','Totals','2022 Average'), as.numeric)
```

```{r echo = FALSE}
MLS %>%
  kbl() %>%
  kable_styling(fixed_thead = T)
```


This next step is to prepare the data just imputed for a future join with
stadium data taken from Wikipedia. Some of the team names from the attendance 
date originating from soccerstadiumdigest.com do not exactly match the way the
names are presented on Wikipedia. In order to be able to join the two data sets
we essentially need to 'standardize' the team names so they match in both data 
sets.

```{r}
# rename points for join - this is to make joining easier in future
# with other data sets 
usl$Team <- gsub("CS Switchbacks FC", "Colorado Springs Switchbacks FC", usl$Team)

MLS$Team <- gsub("LA Galaxy", "Los Angeles Galaxy", MLS$Team)

MLS$Team <- gsub("St. Louis SC", "St. Louis City SC", MLS$Team)

MLS$Team <- gsub("NE Revolution", "New England Revolution", MLS$Team)

MLS$Team <- gsub("LAFC", "Los Angeles FC", MLS$Team)

MLS$Team <- gsub("NYC FC","New York City FC", MLS$Team)

MLS$Team <- gsub("Sporting KC",
                 "Sporting Kansas City", MLS$Team)
MLS$Team <- gsub("SJ Earthquakes",
                 "San Jose Earthquakes", MLS$Team)
MLS$Team <- gsub("NY Red Bulls",
                 "New York Red Bulls", MLS$Team)
usl$Team <- gsub("Rio Grande Valley FC Toros", "Rio Grande Valley FC",
                 usl$Team)
```

Then, using the rbind function we can combine both the USL and MLS attendance
data frames.

```{r}
# join attendance together
leagues_joined <-rbind(usl, MLS)
```

```{r echo=FALSE}
leagues_joined %>%
  kbl() %>%
  kable_styling(fixed_thead = T)
```


## Inputing and Cleaning USL and MLS Stadium Data

Now we are going to input the stadium specific data from Wikipedia.

```{r}
htmlstadium <- read_html("https://en.wikipedia.org/wiki/List_of_Major_League_Soccer_stadiums")

stadium_MLS <- htmlstadium %>% 
  html_node("table.wikitable.sortable") %>% 
  html_table(
    header = TRUE,
  )
```


With some quick manipulations.


```{r}
stadium_MLS <- stadium_MLS %>%
  select(-c(Image,`Ref(s)`)) 

stadium_MLS$Capacity <- readr::parse_number(stadium_MLS$Capacity)

stadium_MLS$Team <- gsub("LA Galaxy", "Los Angeles Galaxy", stadium_MLS$Team)

stadium_MLS$Team <- gsub("St. Louis SC", "St. Louis City SC", stadium_MLS$Team)

stadium_MLS$Team <- gsub("NE Revolution", "New England Revolution", stadium_MLS$Team)

stadium_MLS$Team <- gsub("LAFC", "Los Angeles FC", stadium_MLS$Team)

stadium_MLS$Team <- gsub("NYC FC","New York City FC", stadium_MLS$Team)

stadium_MLS$Team <- gsub("Sporting KC",
                 "Sporting Kansas City", stadium_MLS$Team)
stadium_MLS$Team <- gsub("SJ Earthquakes",
                 "San Jose Earthquakes", stadium_MLS$Team)
stadium_MLS$Team <- gsub("NY Red Bulls",
                 "New York Red Bulls", stadium_MLS$Team)
```


Giving us this table.


```{r echo= FALSE}
stadium_MLS %>%
  kbl() %>%
  kable_styling(fixed_thead = T)
```


The same steps were then taken for the USL Championship stadium data also taken 
from Wikipedia: https://en.wikipedia.org/wiki/2023_USL_Championship_season.

The USL data required an extra cleaning due to some stadium names having "[A]"
by their names, indicating that the stadium was not soccer-specific. Removing
the [A] required the use of the gsub function and a regular expression.


```{r}
html_usl_stadium <- read_html("https://en.wikipedia.org/wiki/2023_USL_Championship_season")

usl_stadium <- html_usl_stadium %>%
  html_node("table.wikitable.sortable") %>%
  html_table(
    header= T
  )
usl_stadium$Capacity <- readr::parse_number(usl_stadium$Capacity)
usl_stadium$Stadium <- gsub("\\Q[A]\\E","",usl_stadium$Stadium)

usl_stadium$Team <- gsub("CS Switchbacks FC", 
                         "Colorado Springs Switchbacks FC", usl_stadium$Team)
usl_stadium$Team <- gsub("Rio Grande Valley FC Toros", "Rio Grande Valley FC",
                 usl_stadium$Team)
```

```{r echo=FALSE}
usl_stadium %>%
  kbl() %>%
  kable_styling(fixed_thead = T)
```


## Joining both the MLS and USL data sets

For this join I utilized the stringdist_inner_join() function from the fuzzyjoin
library, allowing me to join both the attendance and stadium data by matching 
the two tables with the 'team' variable. This is why we went through the 
trouble of renaming some of team names earlier in the process. 


```{r}
# Merge stadium data with leagues data frame
mls_filter <- stadium_MLS %>%
  subset(select = c("Stadium","Team","Capacity"))

stadium_joined <- rbind(mls_filter, usl_stadium)

# joining data for attendance and stadium capacity
leagues_master <- leagues_joined %>%
  stringdist_inner_join(stadium_joined, by ="Team", max_dist = 3) %>%
  select(-Team.y) %>%
  rename(Teams = Team.x)
```


I also created two other columns in the table so we can have insight into what
percentage of the stadium was filled on average for both 2022 and 2023. 


```{r}
# creation of % filled columns
leagues_master <- leagues_master %>%
  mutate("%_2023_Fill" = (Average/Capacity)) %>%
  mutate("%_2022_Fill" = (leagues_master$`2022 Average`/Capacity))
```


Here is the table with the most recent additions:


```{r echo=FALSE}
leagues_master %>%
  kbl() %>%
  kable_styling(fixed_thead = T)
```


I also made separate data frames for both the MLS and USL. This will allow us 
to analyze the leagues separately later on.


```{r}
#MLS DF
mls_master <- MLS %>%
  stringdist_inner_join(stadium_MLS, by ="Team", max_dist = 3) %>%
  select(-Team.y) %>%
  rename(Teams = Team.x)
# creation of % filled columns
mls_master <- mls_master %>%
  mutate("%_2023_Fill" = (Average/Capacity)) %>%
  mutate("%_2022_Fill" = (mls_master$`2022 Average`/Capacity))


#USL DF
usl_master <- usl %>%
  stringdist_inner_join(usl_stadium, by ="Team", max_dist = 3) %>%
  select(-Team.y) %>%
  rename(Teams = Team.x)
# creation of % filled columns
usl_master <- usl_master %>%
  mutate("%_2023_Fill" = (Average/Capacity)) %>%
  mutate("%_2022_Fill" = (usl_master$`2022 Average`/Capacity))

```



# Adding Extra Data Values for Exploratory and Correlative Analysis

Two important variables that can provide insight into attendance are points 
and financials. Points being the point value of the team at the end of the 
season. In professional soccer, teams are awarded 3 points for a win and 1 point
for a draw. Theoretically, teams that have more points would have higher
attendance. 

Furthermore, I also hypothesize that there will be a correlation between
the financial valuation of a clubs roster and it's attendance. Teams that 
have higher payrolls and more star players should be able to draw bigger crowds.

Lastly, I'm also curious about what I call "city moral." If a city has happier 
citizens will they be more likely to support the city's club, because in a sense 
the club represents a city. One way to analyze this is by using approval 
ratings of government officials as a measure of support. Mayoral data is tricky
to find so I decided to use Governor approval ratings. This obviously is not the
most effective strategy so this aspect of the project is really just an extra
step.

Let's now input data for these variables. 


## City and State Values

Firstly, I wanted to add in city/state columns for the main data set that 
has been created thus far. Having the location date is important for joining
Governor approval ratings with the correct teams.

Like in the past steps I decided to scrape data from the USL championship 
Wikipedia page. Because this page has multiple tables I had to take the extra
step of subset a list of the tables from the webpage in order to retrive the
table I needed. This required me to utilize html_nodes rather then html_node.

```{r}
## Adding in city and state data for future analysis ##

html_city_USL <- read_html("https://en.wikipedia.org/wiki/USL_Championship")

usl_city <- html_city_USL %>%
  html_nodes("table.wikitable") %>%
  .[2] %>%
  html_table(header = T)

usl_city <- data.frame(Reduce(rbind, usl_city))

  
# clean table
usl_city<- usl_city[-c(1,14),]
usl_city<- usl_city[,-8]

# separating city and state for USL
usl_city <- separate_wider_delim(usl_city, City, delim = ",", 
                                 names = c("City", "State"))
usl_city <- usl_city %>%
  subset(select = c("Team", "City", "State")) %>%
  rename(Teams = Team)


#adding data to usl specific USL DF
usl_master <- usl_master %>%
  stringdist_inner_join(usl_city, by = "Teams", max_dist = 3) %>%
  select(-Teams.y) %>%
  rename(Teams = Teams.x)
```


Fortunately, my MLS stadium data that was previously retrieved already had
location data. Although I did need to separate the city and state names into
different columns and then add the two variables back into the MLS master sheet.


```{r}
# separating city and state MLS
stadium_MLS <- separate_wider_delim(stadium_MLS, Location, delim = ",",
                                names = c("City", "State"))
stadium_city_mls <- stadium_MLS %>%
  subset(select = c("Team", "City", "State"))%>%
  rename(Teams = Team)

mls_master <- mls_master %>%
  stringdist_inner_join(stadium_city_mls, by = "Teams", max_dist = 3) %>%
  select(-Teams.y) %>%
  rename(Teams = Teams.x)

cities <- rbind(usl_city, stadium_city_mls)


# joining cities with leagues_master DF
leagues_master <- leagues_master %>%
  stringdist_inner_join(cities, by = "Teams", max_dist = 3) %>%
  select(-Teams.y) %>%
  rename(Teams = Teams.x)
```



## 2022 Point Values

Using the same web scraping method as before I utilized data from 
footballdatabase.com for the 2022 MLS points values and data from Wikipedia for 
the 2022 USL Championship point values.


```{r}
##importing team records for MLS and USL then joining into one sheet to join with master ##

#mls
html_MLS_records <- read_html("https://footballdatabase.com/league-scores-tables/united-states-mls-2022")
html_usl_points <- read_html("https://en.wikipedia.org/wiki/2022_USL_Championship_season")

records_MLS <- html_MLS_records %>% 
  html_node("table.table") %>% 
  html_table(
    header = TRUE,
  )

points_MLS <- records_MLS %>%
  select(P, Club) %>%
rename(Teams = Club, '2022 Pts' = P)


#usl
usl_records <- html_usl_points %>%
  html_table(
    header = TRUE
  )

eastern <- purrr::map_dfr(usl_records[5], dplyr::bind_rows)
names(eastern)[2] <- "Teams"
eastern <- eastern %>%
  select(Teams, Pts)

western <- purrr::map_dfr(usl_records[6], dplyr::bind_rows)
names(western)[2] <- "Teams"
western <- western %>%
  select(Teams, Pts)

usl_2022_standings <- rbind(eastern, western) %>%
  rename('2022 Pts' = Pts)

usl_2022_standings$Teams <- gsub("\\Q (C, X)\\E","",usl_2022_standings$Teams)
usl_2022_standings$Teams <- gsub("\\Q Toros\\E"," FC", usl_2022_standings$Teams)

#Removing MLS next pro teams
usl_2022_standings <- usl_2022_standings[- grep("New York Red Bulls II", 
                                                usl_2022_standings$Teams),]
usl_2022_standings <- usl_2022_standings[- grep("Atlanta United 2", 
                                                usl_2022_standings$Teams),]

#joining with both mls and usl masters
usl_master <- usl_master %>%
  stringdist_inner_join(usl_2022_standings, by = "Teams", max_dist = 3)%>%
  select(-Teams.y) %>%
  rename(Teams = Teams.x)

points_MLS$Teams <- gsub("LA Galaxy", "Los Angeles Galaxy", points_MLS$Teams)
mls_master <- mls_master %>%
  stringdist_inner_join(points_MLS, by = "Teams", max_dist = 3)%>%
  select(-Teams.y) %>%
  rename(Teams = Teams.x)


# merging mls and usl points for inner join
Final_standings <- rbind(usl_2022_standings, points_MLS)


# joining final 2022 points with other data 
Final <- leagues_master %>%
  stringdist_inner_join(Final_standings, by = "Teams", max_dist = 3)
Final <- Final[,-13] 
Final <- rename(.data = Final, Teams = Teams.x)
```


Here is the DF with the points variable added.


```{r echo=FALSE}
Final %>%
  kbl() %>%
  kable_styling(fixed_thead = T)
```



## Governor Raitings

```{r include=FALSE}
## loading in other explanatory variable data for regression
governor <- read_csv(here("Data", "Morning Consult Pro Governor Raitings Q2 2023.csv"), 
                     col_names = TRUE )
```


Here is the Governor rating data taken from Morning Consult Pro. This data was
then added to the main data frame.


```{r echo=FALSE}
governor %>%
  kbl() %>%
  kable_styling(fixed_thead = T)
```

```{r include=FALSE}
# joining ratings with main df
ratings <- subset(governor, select = c("Governor", "State", "Approve", "Dissaprove"))

#remove white space for trim
ratings$State <- trimws(ratings$State)
Final$State <- trimws(Final$State)

#join
Final <- left_join(Final, ratings, by = "State")
Final$Approve<- parse_number(x = Final$Approve )
Final$Dissaprove<- parse_number(x = Final$Dissaprove )
```


## Club Finances

This club finance data comes from https://www.transfermarkt.com and is current 
as of October 11, 2023. So, towards the end of the 2023 season. 


```{r}
## loading in club finances (in millions of euros)
finances <- read_xlsx(here("data", "Club Market Values.xlsx"), sheet = "Values")
finances$`Market Value`<- parse_number(finances$`Market Value`)
finances$`Total market value`<- parse_number(finances$`Total market value`)

#joining data
Final <- Final %>%
  stringdist_inner_join(finances, by = "Teams", max_dist = 3)%>%
  select(-Teams.y) %>%
  rename(Teams = Teams.x)


#joining finance with usl master
finances_usl <- read_xlsx(here("Data", "Club Market Values.xlsx"), sheet = "USL")
finances_usl$`Market Value`<- parse_number(finances_usl$`Market Value`)
finances_usl$`Total market value`<- parse_number(finances_usl$`Total market value`)
usl_master <- usl_master %>%
  stringdist_inner_join(finances_usl, by = "Teams", max_dist = 3)%>%
  select(-Teams.y) %>%
  rename(Teams = Teams.x)

#joining finance with mls master
finances_mls <- read_xlsx(here("Data", "Club Market Values.xlsx"), sheet = "MLS")
finances_mls$`Market Value`<- parse_number(finances_mls$`Market Value`)
finances_mls$`Total market value`<- parse_number(finances_mls$`Total market value`)
mls_master <- mls_master %>%
  stringdist_inner_join(finances_mls, by = "Teams", max_dist = 3)%>%
  select(-Teams.y) %>%
  rename(Teams = Teams.x)
```


## 2023 Point Values 

In addition to the 2022 point records I also wanted to include the data from 
2023 for comparison and analysis. This data was importated from a spreadsheet
and joined with the main data set like the past steps in this process. 

```{r include=FALSE}
# mls and usl 2023 points data
#mls
mls_2023_record <- read_xlsx(here("data", "MLS_USL Standings 2023.xlsx"), "MLS")
mls_2023_points <- mls_2023_record %>%
  select(Team, Pts)
mls_2023_points <- mls_2023_points %>%
  rename(Teams = Team)
#usl
usl_2023_record <- read_xlsx(here("data", "MLS_USL Standings 2023.xlsx"), "USL")
usl_2023_points <- usl_2023_record %>%
  select(Team, Pts)
usl_2023_points <- usl_2023_points %>%
  rename(Teams = Team)

# joining 2023 Pts with data frames
mls_master <- mls_master %>%
  stringdist_inner_join(mls_2023_points, by = "Teams", max_dist = 3)%>%
  select(-Teams.y) %>%
  rename(Teams = Teams.x, `Pts 2023` = Pts)

usl_master <- usl_master %>%
  stringdist_inner_join(usl_2023_points, by = "Teams", max_dist = 3)%>%
  select(-Teams.y) %>%
  rename(Teams = Teams.x, `Pts 2023` = Pts)



# adding 2023 points to full data frame with all data
`2023_pts` <- rbind(mls_2023_points, usl_2023_points)

Final <- Final%>%
  stringdist_inner_join(`2023_pts`, by = "Teams", max_dist = 3)%>%
  select(-Teams.y) %>%
  rename(Teams = Teams.x, `Pts 2023` = Pts)
```


With both the 2022 values and 2023 values I created a new column in the data 
frame for the percentage change in values from 2022 to 2023. 

```{r}
# calculating new column based on percentage increase in Pts from 2022 to 2023
  
Final <- Final %>% 
  mutate("Pts_change_yoy" = ((`Pts 2023` -`2022 Pts`)/ `2022 Pts`) * 100)

mls_master <- mls_master %>%
  mutate("Pts_change_yoy" = ((`Pts 2023` -`2022 Pts`)/ `2022 Pts`) * 100)

usl_master <- usl_master %>%
  mutate("Pts_change_yoy" = ((`Pts 2023` -`2022 Pts`)/ `2022 Pts`) * 100)
```



## Quick Data Cleaning

After imputing these explanatory variables I then removed data that
skewed the data. For example one USL teams only had 4 home games, unlike other
teams which played 10 or more games.


```{r}
## cleaning - phoenix and Birmingham only have four games
# number of games are too low
usl_master %>%
  group_by(Games) %>%
  arrange(Games)

usl_master <- usl_master %>% 
  filter(Games >= 10)
Final <- Final %>%
  filter(Games >=10 )


# capacity skews attendance percentage
Final %>%
  filter(`%_2023_Fill`< .10) %>%
  select(Teams, Capacity)
# Miami FC is an outlier and this makes sense given their capacity

Final <- Final %>%
  filter(`%_2023_Fill`> .10)

```


# Exploratoy Plotting and Analysis

```{r echo=FALSE}
## exploratory plotting

# plot of Attendance vs 2022 Points
p <- ggplot(data = Final, mapping = aes(x = `2022 Pts`, y = `%_2022_Fill`)) +
      geom_point(color='blue') + 
  ggtitle(label = "Attendance vs Final Points at End of Season 2022") +
  xlab(label = "Final points in 2022") + 
  ylab(label = "% of stadium filled on average ") + 
  theme(plot.title = element_text(hjust = 0.5), 
        text = element_text(size = 16 ,
                            family = "TT Arial")) +
  scale_y_continuous(labels = scales::percent_format())

p + theme_light() 
```


An initial plot of attendance (as a percentage of the stadium filled) versus the
2022 point values of the various teams across all leagues does not provide any
visual correlation.


Lets look at 2023.

```{r echo=FALSE}
# plot of Attendance vs Pts 2023
q <- ggplot(data = Final, mapping = aes(x = `Pts 2023`, y = `%_2023_Fill`)) +
  geom_point(color='blue') + 
  ggtitle(label = "Attendance vs Final Points at End of Season 2023") +
  xlab(label = "Final points in 2023") + 
  ylab(label = "% of stadium filled on average ") + 
  theme(plot.title = element_text(hjust = 0.5), 
        text = element_text(size = 16 ,
                            family = "TT Arial")) +
  scale_y_continuous(labels = scales::percent_format())
q + theme_light() 
```


Much of the same result.


```{r echo=FALSE}
# bar chart of capacity across leagues
ggplot(data = Final, aes(x = Teams, y = Capacity)) +
  geom_col(aes(fill = Capacity)) + theme(plot.title = element_text(hjust = 0.5),
                     axis.title.x=element_blank(),
                     axis.text.x=element_blank(),
                     axis.ticks.x=element_blank()) +
  labs(title = "Capacity of US Soccer Venues") + 
  scale_fill_gradient(low = "blue",
                      high = "red",
                      space = "Lab")
```

Looking as the capacity of stadiums across both leagues we see a wide difference
in size.

```{r}
summary(Final$Capacity)
```


Maybe the capacity of the stadium has some kind of effect on the ability for 
teams to fill the stadium? 

```{r echo=FALSE}
# 2022 Attendance vs capacity
ggplot(data = Final, aes(x = Capacity, y = `%_2022_Fill`)) +
  geom_point() + theme_clean() +
  scale_y_continuous(labels = scales::percent_format())
```

There does appear to be a slight trend that bigger stadiums are actually 
more well filled. Perhaps because clubs with bigger stadiums are more well
established and popular. 

Looking at governor support versus attendance there does not appear to be any 
real correlation between the two variables. 

```{r echo=FALSE }
# attendance vs governor approval
ggplot(data = Final, aes(x = Approve, y = `%_2023_Fill`)) +
  geom_point() + theme_few()+
  scale_y_continuous(labels = scales::percent_format())+
  scale_x_continuous(labels = scales::percent_format(scale = 1))+
  ggtitle(label = "Attendance vs Governor Approval")
```



```{r echo=FALSE}
ggplot(data = mls_master %>% filter(`Total market value` < 70), 
       aes(x = `Total market value`, y = `%_2023_Fill`)) +
  geom_point(color = "purple") +
  ggtitle(label = "Attendence vs Market Value in MLS 2023") +
  theme_clean()+ scale_y_continuous(labels = scales::percent_format())+
  theme(plot.title = element_text(hjust = 0.5)) + 
  xlab(label = "Total Market Value (Millions of Euros)") + 
  ylab(label = "% Of Stadium Filled on Average ") + 
  geom_smooth(method = "lm", se = FALSE)
```

This plot of attendance versus market value indicates that there is potentially
a correlation between the two variables. 


However, the USL data does not indicate the same correlation.

```{r echo=FALSE}
ggplot(data = usl_master, aes(x = `Total market value`, y = `%_2023_Fill`)) +
  geom_point(color = "purple") +
  ggtitle(label = "Attendence vs Market Value in USL 2023") +
  theme_clean()+ scale_y_continuous(labels = scales::percent_format())+
  theme(plot.title = element_text(hjust = 0.5)) + 
  xlab(label = "Total Market Value (Millions of Euros)") + 
  ylab(label = "% Of Stadium Filled on Average ")
```

This chart below illustrates the stark differences between the two leagues. All 
the USL teams have a significantly lower value and little difference between 
the size of values. 

```{r echo=FALSE}
# illustrates that leagues need to largely be treated separately
ggplot(data = Final, aes(x = `Total market value`, y = `%_2023_Fill`)) +
  geom_point(color = "skyblue") +
  ggtitle(label = "Attendence vs Market Value in US soccer") +
  theme_clean()+ scale_y_continuous(labels = scales::percent_format())+
  theme(plot.title = element_text(hjust = 0.5)) + 
  xlab(label = "Total Market Value (Millions of Euros)") + 
  ylab(label = "% Of Stadium Filled on Average ")

```

Looking further at the relation between season-ending point values and 
Attendance. We can see that there is potentially some correlation within the
MLS data set for the 2023 season. 

```{r echo=FALSE}
# focusing on MLS specifically
ggplot(data = mls_master, aes(x = `Pts 2023`, y = `%_2023_Fill`)) +
  geom_point(color = "black") +
  ggtitle(label = "Attendence vs Market Value in MLS") +
  scale_y_continuous(labels = scales::percent_format())+
  theme(plot.title = element_text(hjust = 0.5), panel.background 
        = element_rect(fill = "skyblue")) + 
  xlab(label = "2023 Total Pts") + 
  ylab(label = "% Of Stadium Filled on Average")
```


R can confirm this with the cor function,

```{r}
cor(mls_master$`Pts 2023`, mls_master$`%_2023_Fill`)
```



# Regression Analysis

Using the prior exploratory analysis as a guide it appears that there
are potentially some correlative relationships between attendance rates and 
market valuations, as well as season-ending point values. 

## First Model (MLS)

```{r}
#MLS models
# removing the Messi effect inflating market value with filter
model_mls <- lm(`%_2023_Fill` ~ `Total market value`,mls_master %>% 
                  filter(`Total market value` < 70))
summary(model_mls)
```

This first model does indicate a some relationship between Total Market Value
and % of the stadium filled in 2023 for the MLS data set. Although the adjusted 
R-squared on 0.14 indicates weak correlation. 

## Second Model (MLS) - YOY as Response Variable 

However, a model which incorporates year-over-year attendance rate as the
response variable and the 2022 season ending point and total market value 
as the explanatory variables offers a stronger fit.

```{r}
# change in response variable to YOY changes in attendance
model_mls2 <- lm(`YOY` ~ `2022 Pts` + `Total market value`, mls_master)
summary(model_mls2)
```

The Adjusted R-squared of 0.3286 is a much stronger fit and Total Market Value
has a high level of significance. 


## Third Model (USL) - YOY as Response Variable

The same model as above, except it utilizes USL data, indicates the same
relationship and fit as the model utilizing MLS data. 

```{r}
model_usl2 <- lm(`YOY` ~ `2022 Pts` + `Total market value`, mls_master)
summary(model_usl2)
```


# Conclusions

- Correlation between market value and attendance. Higher market value
correlates with better attendance year-over-year, controlling for the overall
success of the team (measured by standings points).

- Governor approval does not appear to be correlated with sport attendance.

- Stadium size is not a primary driver of attendance.

- Future investigations might look at median income and overall population size.



# Appendix

A look at the data at a regional level.

```{r include=FALSE}
# regional data to look for regional trends

regions <- read_csv(here("data","us census bureau regions and divisions.csv"),
                    col_names = TRUE)

# joining data with other means
by_region_usa <- Final %>%
  inner_join(regions, by = "State")
```

```{r echo=FALSE}
region_mean <- by_region_usa %>%
  group_by(Region) %>% 
  summarise(meanPercentageFill = mean(`%_2023_Fill`, na.rm = TRUE), 
            medianPercentageFill = median(`%_2023_Fill`, na.rm = TRUE))

a <- ggplot(region_mean, aes(x = Region, y = meanPercentageFill))

a + geom_col(fill = "Sky Blue")+
  ggtitle(label = "Percentage of Stadium Filled on Average in 2023 by Region") +
  xlab(label = "Region") + 
  ylab(label = "Percentage of Stadium Filed on Avreage") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_clean()
```

A quick look at the data indicates that the Midwest and the Northeast
show the most consistent support.


